SDG contributions

### 2023-09-15 article/article.qmd
add OA figures

@@

# Institutional prestige
```{r}
papers_per_affiliation_per_fos <- papers %>% 
  left_join(author_paper_affiliations) %>% 
  left_join(affils) %>% 
  group_by(authorid, paperid) %>% 
  mutate(frac_value = 1/n()) %>% 
  group_by(affiliationid, year, SDG_label) %>% 
  summarise(n_frac_papers = sum(frac_value, na.rm = TRUE),
            n_frac_citations = sum(frac_value * citations_norm, na.rm = TRUE)) %>% 
  collect()
```

@@

papers_per_affiliation_per_w_leiden <- papers_per_affiliation_per_fos %>%
  mutate(affiliationid = as.numeric(affiliationid),
         year = as.character(year)) %>% # needed for merging
  left_join(affil_leiden_key) %>%
  left_join(leiden_small_local, by = join_by(University, 
                                             year == last_year_of_period))
```

## Inclusion into Leiden
```{r inclusion_in_leiden, fig.width=9, fig.height=5}
papers_per_affiliation_per_w_leiden %>% 
  filter(year == "2018") %>% 
  mutate(in_leiden = if_else(!is.na(University), "matched", "unmatched")) %>% 
  pivot_longer(starts_with("n_"), names_to = "indicator") %>% 
  mutate(indicator = if_else(
    indicator == "n_frac_citations", 
    "# of citations (fractional)",
    "# of publications (fractional)"
  )) %>% 
  ggplot(aes(SDG_label, value, fill = in_leiden)) +
  facet_wrap(vars(indicator)) + 
  geom_boxplot() +
  scale_y_log10(labels = scales::comma) +
  colorspace::scale_fill_discrete_qualitative() +
  labs(x = NULL, y = NULL, fill = "Leiden Ranking",
       title = "Difference according to mapping to Leiden ranking") +
  theme(legend.position = "top")
```

@@

papers_per_affiliation_per_w_leiden %>%
  filter(year == 2018) %>% 
  plot_bivariate(P_top10, 800, 20000) +
  labs(x = "# of publications of University that is in top 10% of citations 2015-2018 (fractional)",
       y = "# of papers towards SDG in 2018 (fractional)",
       title = "# of top 10% cited publications vs. SDG production")
```

## Over time
```{r}
pdata <- papers_per_affiliation_per_w_leiden %>% 
  filter(!is.na(P_top10)) %>% 
  group_by(Period) %>% 
  mutate(across(c(P_top10, PP_top10, impact_P), cut_quantiles))

@@

## Baseline comparison
```{r}
mag_2021_papers_per_affiliation_per_fos <- mag_2021_papers %>% 
  left_join(mag_2021_paper_author_affil) %>% 
  group_by(authorid, paperid) %>% 
  mutate(frac_value = 1/n()) %>% 
  group_by(affiliationid, year) %>% 
  summarise(n_frac_papers = sum(frac_value, na.rm = TRUE)) %>% 
  filter(!is.na(affiliationid)) %>% 
  collect()

# match with leiden data
mag_2021_papers_per_affiliation_per_w_leiden <- mag_2021_papers_per_affiliation_per_fos %>%
  mutate(affiliationid = as.numeric(affiliationid)) %>% # needed for merging
  left_join(affil_leiden_key) %>%
  left_join(leiden_small_local, by = join_by(University, 
                                             year == last_year_of_period))

# calculate baseline data to fit current plot
baseline_data <- mag_2021_papers_per_affiliation_per_w_leiden %>% 
  filter(!is.na(P_top10)) %>% 
  group_by(Period) %>% 
  mutate(across(c(P_top10, PP_top10, impact_P), cut_quantiles)) %>% 
  group_by(year, P_top10) %>% 
  summarise(n = sum(n_frac_papers, na.rm = TRUE)) %>% 
  group_by(year) %>% 
  mutate(prop = n/sum(n))
```

@@

# Open Access

>>>>

# World Bank data
wb_indicators <- spark_read_csv(
  sc,
  "/tklebel/SDG/world_bank_indicators.csv",
  name = "wb_indicators")

wb_countries <- read_csv(here::here("data/external/WDICountry.csv"))

@@

# Institutional prestige
```{r}
papers_per_affiliation_per_sdg <- papers %>% 
  left_join(author_paper_affiliations) %>% 
  left_join(affils) %>% 
  group_by(authorid, paperid) %>% 
  mutate(frac_value = 1/n()) %>% 
  group_by(affiliationid, year, SDG_label) %>% 
  summarise(n_frac_papers = sum(frac_value, na.rm = TRUE),
            n_frac_oa_papers = sum(frac_value * as.numeric(is_oa), na.rm = TRUE),
            n_frac_citations = sum(frac_value * citations_norm, na.rm = TRUE)) %>% 
  collect()
```

@@

papers_per_affiliation_per_SDG_w_leiden <- papers_per_affiliation_per_sdg %>%
  mutate(affiliationid = as.numeric(affiliationid),
         year = as.character(year)) %>% # needed for merging
  left_join(affil_leiden_key) %>%
  left_join(leiden_small_local, by = join_by(University, 
                                             year == last_year_of_period))
```

## Inclusion into Leiden
```{r inclusion_in_leiden, fig.width=9, fig.height=5}
papers_per_affiliation_per_SDG_w_leiden %>% 
  filter(year == "2018") %>% 
  mutate(in_leiden = if_else(!is.na(University), "matched", "unmatched")) %>% 
  pivot_longer(starts_with("n_"), names_to = "indicator") %>% 
  mutate(indicator = if_else(
    indicator == "n_frac_citations", 
    "# of citations (fractional)",
    "# of publications (fractional)"
  )) %>% 
  ggplot(aes(SDG_label, value, fill = in_leiden)) +
  facet_wrap(vars(indicator)) + 
  geom_boxplot() +
  scale_y_log10(labels = scales::comma) +
  colorspace::scale_fill_discrete_qualitative() +
  labs(x = NULL, y = NULL, fill = "Leiden Ranking",
       title = "Difference according to mapping to Leiden ranking") +
  theme(legend.position = "top")
```

@@

papers_per_affiliation_per_SDG_w_leiden %>%
  filter(year == 2018) %>% 
  plot_bivariate(P_top10, 800, 20000) +
  labs(x = "# of publications of University that is in top 10% of citations 2015-2018 (fractional)",
       y = "# of papers towards SDG in 2018 (fractional)",
       title = "# of top 10% cited publications vs. SDG production")
```

## Over time
```{r}
pdata <- papers_per_affiliation_per_SDG_w_leiden %>% 
  filter(!is.na(P_top10)) %>% 
  group_by(Period) %>% 
  mutate(across(c(P_top10, PP_top10, impact_P), cut_quantiles))

@@

## Baseline comparison
```{r}
mag_2021_papers_per_affiliation_per_sdg <- mag_2021_papers %>% 
  left_join(mag_2021_paper_author_affil) %>% 
  group_by(authorid, paperid) %>% 
  mutate(frac_value = 1/n()) %>% 
  group_by(affiliationid, year) %>% 
  summarise(n_frac_papers = sum(frac_value, na.rm = TRUE)) %>% 
  filter(!is.na(affiliationid)) %>% 
  collect()

# match with leiden data
mag_2021_papers_per_affiliation_per_SDG_w_leiden <- mag_2021_papers_per_affiliation_per_sdg %>%
  mutate(affiliationid = as.numeric(affiliationid)) %>% # needed for merging
  left_join(affil_leiden_key) %>%
  left_join(leiden_small_local, by = join_by(University, 
                                             year == last_year_of_period))

# calculate baseline data to fit current plot
baseline_data <- mag_2021_papers_per_affiliation_per_SDG_w_leiden %>% 
  filter(!is.na(P_top10)) %>% 
  group_by(Period) %>% 
  mutate(across(c(P_top10, PP_top10, impact_P), cut_quantiles)) %>% 
  group_by(year, P_top10) %>% 
  summarise(n = sum(n_frac_papers, na.rm = TRUE)) %>% 
  group_by(year) %>% 
  mutate(prop = n/sum(n))
```

@@

# Open Access
```{r}
# aggregate oa_status -----
oa_status <- papers %>%
  filter(!is.na(is_oa)) %>%
  select(paperid, SDG_label, year, is_oa, provider_cat)

oa_per_year <- oa_status %>%
  count(SDG_label, year, is_oa) %>%
  collect()
```

```{r, sdg_oa_by_fos}
oa_per_year %>%
  group_by(SDG_label, year) %>%
  mutate(oa_share = n/sum(n)) %>%
  filter(is_oa) %>%
  ggplot(aes(as_year(year), oa_share, colour = fix_sdg(SDG_label),
             group = fix_sdg(SDG_label))) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
  colorspace::scale_color_discrete_qualitative() +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme(legend.position = "top")
```

## By country
```{r}
oa_per_affiliation <- oa_status %>%
  left_join(author_paper_affiliations) %>%
  left_join(affils)

oa_per_affiliation_selected <- oa_per_affiliation %>%
  group_by(paperid) %>%
  mutate(frac_count = 1 / max(authorsequencenumber, na.rm = TRUE)) %>%
  select(paperid, authorid, is_oa, provider_cat, year, country, frac_count,
         SDG_label)

oa_per_country <- oa_per_affiliation_selected %>%
  filter(year >= 2015 & year <= 2018) %>% 
  group_by(country, is_oa) %>%
  summarise(sum_frac_oa = sum(frac_count)) %>%
  mutate(prop_oa = sum_frac_oa/sum(sum_frac_oa),
         sum_frac_total = sum(sum_frac_oa)) %>%
  collect()
```

```{r oa_sdg_per_country_gdp_p_cap, fig.width=9}
wb_local <- wb_indicators %>%
  filter(year >= 2015 & year <= 2018) %>%
  group_by(country_name, country_code, indicator_code, indicator_name) %>% 
  summarise(value = mean(value, na.rm = TRUE)) %>% 
  collect()

proper_countries <- wb_countries %>% 
  filter(!is.na(`Currency Unit`)) %>% 
  select(country_code = `Country Code`, short_name = `Short Name`,
         region = Region, income_group = `Income Group`)

oa_with_gdp_per_cap <- oa_per_country %>%
  left_join(wb_local, by = c("country" = "country_code")) %>%
  select(-indicator_name) %>%
  filter(indicator_code %in% c("NY.GDP.PCAP.KD")) %>%
  pivot_wider(names_from = indicator_code, values_from = value) %>%
  drop_na() %>%
  filter(is_oa)

p <- oa_with_gdp_per_cap %>%
  filter(sum_frac_total >= 50) %>%
  left_join(proper_countries, by = c("country" = "country_code")) %>% 
  ggplot(aes(NY.GDP.PCAP.KD, prop_oa)) +
  geom_point(aes(size = sum_frac_total)) +
  labs(x = "GDP per capita", y = "% of publications which are OA",
       size = "# of publications") +
  scale_size_continuous(trans = "sqrt", labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_log10(breaks = c(1e+03, 2e+03, 5e+03, 1e+04, 2e+04, 5e+04, 1e+05),
                labels = function(x) scales::comma(x, prefix = "$")) +
  theme_bw()

p1 <- p +
  aes(colour = region)

p1 +
    geom_smooth(aes(colour = NULL), alpha = .3, show.legend = FALSE,
                colour = "grey30") +
  labs(colour = "World region")
```
### split by SDG
```{r}
oa_per_country_per_SDG <- oa_per_affiliation_selected %>%
  filter(year >= 2015 & year <= 2018) %>% 
  group_by(country, SDG_label, is_oa) %>%
  summarise(sum_frac_oa = sum(frac_count)) %>%
  mutate(prop_oa = sum_frac_oa/sum(sum_frac_oa),
         sum_frac_total = sum(sum_frac_oa)) %>%
  collect()

oa_sdg_with_gdp_per_cap <- oa_per_country_per_SDG %>%
  left_join(wb_local, by = c("country" = "country_code")) %>%
  select(-indicator_name) %>%
  filter(indicator_code %in% c("NY.GDP.PCAP.KD")) %>%
  pivot_wider(names_from = indicator_code, values_from = value) %>%
  drop_na() %>%
  filter(is_oa)
```

```{r oa_by_sdg, fig.height=7, fig.width=9}
oa_sdg_with_gdp_per_cap %>%
  filter(sum_frac_total >= 50) %>%
  left_join(proper_countries, by = c("country" = "country_code")) %>% 
  ggplot(aes(NY.GDP.PCAP.KD, prop_oa)) +
  geom_point(aes(size = sum_frac_total, colour = region)) +
  geom_smooth(alpha = .3, show.legend = FALSE, colour = "grey30") +
  labs(x = "GDP per capita", y = "% of publications which are OA",
       colour = "World region", size = "# of publications") +
  scale_size_continuous(trans = "sqrt", labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_log10(breaks = c(1e+03, 2e+03, 5e+03, 1e+04, 2e+04, 5e+04, 1e+05),
                labels = function(x) scales::comma(x, prefix = "$")) +
  theme_bw() +
  facet_wrap(vars(fct_relevel(SDG_label, "SDG_13", after = 3)), nrow = 2) +
  theme(legend.position = c(.8, .2))
```

### Correlate with institutional prestige
```{r}
affil_oa <- papers_per_affiliation_per_SDG_w_leiden %>% 
  mutate(oa_share = n_frac_oa_papers / n_frac_papers)
```

```{r oa_prestige_correlation, fig.width=8, fig.height=4}
affil_oa %>% 
  group_by(year, SDG_label) %>% 
  filter(n_frac_papers > 50) %>% 
  summarise(cor = cor(oa_share, P_top10, use = "pairwise.complete.obs")) %>% 
  ggplot(aes(year, cor, colour = fix_sdg(SDG_label))) +
  geom_line() +
  geom_point() +
  colorspace::scale_color_discrete_qualitative() +
  scale_y_continuous(limits = c(0, .6)) +
  labs(y = expression(paste("Correlation between OA shares in SDG & P"["top 10%"])),
       x = NULL, colour = NULL) +
  theme(legend.position = "top")
```
